Reasonably Programmable Literal Notation

### 2018-06-22 icfp18/syntax-icfp18.tex
edits through 2.2.4

\begin{figure}
\begin{lstlisting}[deletekeywords={spliced}]
  module Segment : { 
    type t = {start_pos: int, end_pos: int};
    let read_to : string => Lexing.lexbuf => Segment.t; };
  exception ExpansionError({msg: string, loc: option(Segment.t)});
  module ProtoExpr : { 
    type t = Parsetree.expression; /* from the OCaml compiler library */
    let spliced : Segment.t => Parsetree.core_type => t; };
  notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
  notation $proto_expr at ProtoExpr.t = $re_expr;
\end{lstlisting}
\vspace{-2px}
\caption{A fragment of the \li{Relit} module signature relevant to expression TLM providers.}
\label{fig:relit-util}
\vspace{-5px}
\end{figure}
    % let match_spliced : t => option(Segment.t, ProtoTyp.t);

@@

\subsubsection{Correctness Criteria}
\label{sec:correctness-criteria}
In order to maintain the abstract reasoning principles discussed in Sec. \ref{sec:client-perspective}, the system \emph{validates} each proto-expansion generated by the parser as follows:
\begin{enumerate}
  \item First, the segmentation is computed and validated. This involves checking two criteria:
    \begin{enumerate}
      \item The segments must be in-bounds, of positive extent, non-overlapping and separated by at least one character.
      \item The segment types must encode valid types assuming only the TLM dependencies.
    \end{enumerate}
  \item Second, the proto-expansion is typechecked assuming only the TLM dependencies (plus opened \li{Pervasives}), treating each splice reference as a variable of the specified segment type.
\end{enumerate}
If proto-expansion validation fails, the client is notified that the applied TLM is incorrectly implemented. The proto-expansion validation criteria are exactly the correctness criteria that parser writers must consider, quantifying over all possible literal bodies for which a proto-expansion is generated, in order to avoid exposing this class of compile-time error to clients.

\subsubsection{MetaOCaml}\label{sec:metaocaml} In ``vanilla'' OCaml, it is not possible to express a datatype that classifies only valid proto-expansions according to the correctness criteria above. The MetaOCaml type system comes closer by extending the semantics of OCaml with values of type \li{'a code}, which are constructed using a primitive typed quasiquotation operation \cite{DBLP:conf/flops/Kiselyov14,taha2004gentle}. In MetaOCaml-BER, which is an active implementation of (an extension of) MetaOCaml as a fork of the compiler, and \li{ppx_stage}, which implements a subset of MetaOCaml as a compiler plugin \cite{ppxstage}, a value of type \li{'a code} is represented as, and can be coerced safely to, an OCaml parse tree for the corresponding expression of type \li{'a} \cite{DBLP:conf/flops/Kiselyov14}. It is therefore possible to implement a TLM parser using \li{'a code} values internally, doing the coercion only at the end. However, the typing guarantee is relative to the context where the quotation was constructed, i.e. the parser's implementation, so a validation step is still needed to ensure that the assumptions are consistent with those specified by the \li{dependencies} clause.

The segmentation criteria cannot be statically enforced and, in both mentioned implementations of MetaOCaml, it is impossible to express splice references directly because they do not yet support terms that contain explicit type annotations. There are various workarounds, e.g. by defining for each possible segment type \li{t} a dummy value of type \li{t code} and then post-processing the parse tree after the coercion step. 
Overall, however, this approach can substantially increase a TLM provider's confidence in the type correctness of the expansion logic, with the caveats just mentioned. With \li{ppx_stage}, it is possible to compile the parser separately without requiring that clients use the plugin. 
TLMs can be applied inside quoted code as long as the applied TLM is accessed through a dependency of the TLM being defined (so that internal dependency references correctly resolve). %MetaOCaml-BER does not change the parse tree or bytecode representation of the OCaml system, and values of type \li{'a code} are not part of the public interface of a TLM parser, so it should, in principle, be possible to compile a parser using MetaOCaml-BER for use in  ``vanilla'' OCaml (and therefore Reason) programs. However, MetaOCaml-BER does not suitably package the necessary pieces of its run-time system, so we have not verified this conjecture.

@@

Much of the research on macro systems has been for languages in the LISP tradition \cite{mccarthy1978history} that do not have rich static type structure. The formal macro calculus studied by \citet{DBLP:conf/esop/HermanW08} (which is not capable of expressing new literal forms, for the reason just discussed) uses types only as a technical advice in reasoning about the binding structure of the generated expansion. The Scala macro system does support reasoning abstractly about \textbf{Expansion Typing} due to the return type annotations. The full calculus we have defined is the first detailed type-theoretic account of a typed, hygienic macro system of any design for an ML- or Scala-like language, i.e. one with a rich static type structure and support for pattern matching. Many of the basic mechanisms would be relevant even if support for parsing literal bodies was removed and replaced with term rewriting. Segmentations can be considered a refinement of the tree paths in prior work \cite{Herman10:Theory,gorn1965explicit}.

@@

At several points in the paper, we allude to editor integration. However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered. Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor \cite{DBLP:conf/popl/OmarVHAH17}.

>>>>

\begin{figure}
\begin{lstlisting}[deletekeywords={spliced}]
  module Segment : { 
    type t = {start_pos: int, end_pos: int};
    let read_to : string => Lexing.lexbuf => Segment.t; };
  exception ExpansionError({msg: string, loc: option(Segment.t)});
  module ProtoExpr : { 
    type t = Parsetree.expression; /* from the OCaml compiler library */
    let spliced : Segment.t => Parsetree.core_type => t; };
  notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
  notation $proto_expr at ProtoExpr.t = $re_expr;
\end{lstlisting}
\vspace{-2px}
\caption{A fragment of the \li{Relit} module's signature relevant to expression TLM providers.}
\label{fig:relit-util}
\vspace{-5px}
\end{figure}
    % let match_spliced : t => option(Segment.t, ProtoTyp.t);

@@

\subsubsection{Correctness Criteria}
\label{sec:correctness-criteria}
In order to maintain the abstract reasoning principles discussed in Sec. \ref{sec:client-perspective}, the system \emph{validates} each proto-expansion generated by the parser as follows:
\begin{enumerate}
  \item First, the segmentation is computed and validated. This involves checking two criteria:
    \begin{enumerate}
      \item The segments must be in-bounds, of positive extent, non-overlapping and separated by at least one character.
      \item The segment types must encode valid types assuming only the TLM dependencies.
    \end{enumerate}
  \item Second, the proto-expansion is typechecked assuming only the TLM dependencies (plus opened \li{Pervasives}), treating each splice reference as a variable of the specified segment type.
\end{enumerate}
If validation fails, the client is notified that the applied TLM is incorrectly implemented.% The proto-expansion validation criteria are exactly the correctness criteria that parser writers must consider, quantifying over all possible literal bodies for which a proto-expansion is generated, in order to avoid exposing this class of compile-time error to clients.

\subsubsection{MetaOCaml}\label{sec:metaocaml} In ``vanilla'' OCaml, it is not possible to define a datatype that classifies only valid proto-expansions according to the correctness criteria above. The MetaOCaml type system comes closer by extending the semantics of OCaml with values of type \li{'a code}, which are constructed using a primitive typed quasiquotation operation \cite{DBLP:conf/flops/Kiselyov14,taha2004gentle}. In MetaOCaml-BER, which is an active implementation of (an extension of) MetaOCaml as a fork of the compiler, and \li{ppx_stage}, which implements a subset of MetaOCaml as an ordinary compiler plugin rather than a fork of the compiler \cite{ppxstage}, a value of type \li{'a code} is represented as, and can be coerced safely to, an OCaml parse tree for the corresponding expression of type \li{'a} \cite{DBLP:conf/flops/Kiselyov14}. It is therefore possible to implement a TLM parser using \li{'a code} values internally, doing the coercion only at the end. However, the typing guarantee is relative to the context where the quotation was constructed, i.e. the parser's implementation, so a validation step is still needed to ensure that the assumptions are consistent with those specified by the \li{dependencies} clause.

The segmentation criteria cannot be statically enforced and, in both mentioned implementations of MetaOCaml, it is impossible to express splice references directly because they do not yet support terms that contain explicit type annotations. There are various workarounds, e.g. by defining for each possible segment type \li{t} a dummy value of type \li{t code} and then post-processing the parse tree after the coercion step.

Overall, this approach can substantially increase a TLM provider's confidence in the type correctness of the expansion logic, with the caveats just mentioned. With \li{ppx_stage}, it is possible to compile the parser separately without requiring that clients use the plugin. 
TLMs can be applied inside quoted code as long as the applied TLM is accessed through a dependency of the TLM being defined (so that internal dependency references correctly resolve). %MetaOCaml-BER does not change the parse tree or bytecode representation of the OCaml system, and values of type \li{'a code} are not part of the public interface of a TLM parser, so it should, in principle, be possible to compile a parser using MetaOCaml-BER for use in  ``vanilla'' OCaml (and therefore Reason) programs. However, MetaOCaml-BER does not suitably package the necessary pieces of its run-time system, so we have not verified this conjecture.

@@

Much of the research on macro systems has been for languages in the LISP tradition \cite{mccarthy1978history} that do not have rich static type structure. The formal macro calculus studied by \citet{DBLP:conf/esop/HermanW08} (which is not capable of expressing new literal forms, for the reason just discussed) uses types only as a technical advice in reasoning about the binding structure of the generated expansion. The Scala macro system does support reasoning abstractly about \textbf{Expansion Typing} due to the return type annotations. The full calculus we have defined is the first detailed type-theoretic account of a typed, hygienic macro system of any design for an ML- or Scala-like language, i.e. one with a rich static type structure and support for pattern matching. Many of the mechanisms would be relevant even if support for parsing literal bodies was removed and replaced with term rewriting. Segmentations can be considered a refinement of the tree paths in prior work \cite{Herman10:Theory,gorn1965explicit}.

@@

At several points, we allude to editor integration. However, several important questions having to do with TLM-specific syntax highlighting, incremental parsing and error recovery \cite{graham1979practical} remain to be considered. Another interesting direction would be to generalize TLMs to support non-textual notation in the setting of a structure editor \cite{DBLP:conf/popl/OmarVHAH17}.
